`traveltime`: an R package to calculate walking and motorised travel time from user-specified locations

### 2025-01-24 paper/paper.qmd
fixes cross refs - resolve #43 - did I create this issue recently when it wasn't previously a problem? :shrug:

Understanding and mapping the time to travel among locations is useful for many activities from urban planning [@zahavi1974traveltime] to public health [@weiss2020global] and likely a myriad others we haven't thought of. Here we present a software package --- `traveltime` --- written in and for the language R [@Rref]. `traveltime` enables a user to supply a set of geographic point locations and an extent of interest, and calculate the motorised or walking travel time over that extent. The result is a raster of 30 arcsecond resolution (approximately 0.008333 decimal degrees, or just below 1 km$^2$ at the equator) where the value in each cell is the lowest travel time in minutes to any of the supplied point locations over the extent of interest.

@@

Reference documents for `traveltime` are available at <https://idem-lab.github.io/traveltime/>. Although this article is intended to be the key reference for the `traveltime` package, we suggest citations of the package should be accompanied by citing the underlying work by @weiss2018global; @weiss2020global.

@@

Global maps of travel time to cities [@weiss2018global] and health care facilities [@weiss2020global] `@fig-tt_healthcare` have generated significant interest and use. Collectively @weiss2018global ; @weiss2020global have over 1300 citations (Google Scholar, at 10th December 2024), and the city data set is available to R users through the widely-used `geodata` package [@geodata]. There is clear demand for these type of products.

@@

We can pass in our basemap `sin` @fig-basemap, a `SpatVector`, directly as the `extent`. It would also be possible pass in a `SpatRaster` with extent equal to our area of interest, or specify coordinates either as a vector, matrix, or `SpatExtent`.

@@

Taking a look at these data, below we plot the friction surface raster `friction_singapore`, with the vector boundary of `sin` as a dashed grey line, and `stations` as grey points `@fig-data`. Higher values of friction indicate more time travelling across a given cell.

@@

We present those results in `@fig-result`:

@@

Note that the above raster include infinite (`Inf`) values. In `@fig-data`, a number of islands to the south are shown as cells unconnected with the mainland. These raster cells for these islands appear absent in `@fig-result`. Because they are not connected, the calculated travel time is infinite, and so these cells do not appear in the plot `@fig-result`.

>>>>

Understanding and mapping the time to travel among locations is useful for many activities from urban planning [@zahavi1974traveltime] to public health [@weiss2020global] (@ref-tt_healthcare) and likely a myriad others we haven't thought of. Here we present a software package --- `traveltime` --- written in and for the language R [@Rref]. `traveltime` enables a user to supply a set of geographic point locations and an extent of interest, and calculate the motorised or walking travel time over that extent. The result is a raster of 30 arcsecond resolution (approximately 0.008333 decimal degrees, or just below 1 km$^2$ at the equator) where the value in each cell is the lowest travel time in minutes to any of the supplied point locations over the extent of interest.

@@

`traveltime` is available from [R-Universe](https://idem-lab.r-universe.dev/traveltime) and [GitHub](https://github.com/idem-lab/traveltime), and reference documents at <https://idem-lab.github.io/traveltime/>. Although this article is intended to be the key reference for the `traveltime` package, we suggest citations of the package should be accompanied by citing the underlying work by @weiss2018global; @weiss2020global.

@@

Global maps of travel time to cities [@weiss2018global] and health care facilities [@weiss2020global] (@fig-tt_healthcare) have generated significant interest and use. Collectively @weiss2018global ; @weiss2020global have over 1300 citations (Google Scholar, at 10th December 2024), and the city data set is available to R users through the widely-used `geodata` package [@geodata]. There is clear demand for these type of products.

@@

We can pass in our basemap `sin` (@fig-basemap), a `SpatVector`, directly as the `extent`. It would also be possible pass in a `SpatRaster` with extent equal to our area of interest, or specify coordinates either as a vector, matrix, or `SpatExtent`.

@@

Taking a look at these data, below we plot the friction surface raster `friction_singapore`, with the vector boundary of `sin` as a dashed grey line, and `stations` as grey points @fig-data. Higher values of friction indicate more time travelling across a given cell.

@@

We present those results in @fig-result:

@@

Note that the above raster include infinite (`Inf`) values. In @fig-data, a number of islands to the south are shown as cells unconnected with the mainland. These raster cells for these islands appear absent in @fig-result. Because they are not connected, the calculated travel time is infinite, and so these cells do not appear in the plot @fig-result.
